---
title: "Introduction to sfclust"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sfclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `sfclust` package provides a Bayesian framework for spatial functional clustering. It
is designed to analyze spatially distributed functional data and discover meaningful
clusters. The package employs advanced statistical methods to integrate spatial and
functional information, making it applicable in diverse domains such as environmental
studies, crime analysis, and epidemiology. For more information, refer to the detailed
methodology and case studies included in the package documentation.

# Installation

To install the development version of `sfclust` from GitHub, use the following commands:

```{r eval=FALSE}
# Install development version from GitHub
devtools::install_github("RuimanZhong/sfclust")
```

Additionally, the `sfclust` package relies on the `INLA` package for efficient Bayesian
computation. You can install it as follows:

```{r eval=FALSE}
# Install INLA
install.packages("INLA", repos = c(getOption("repos"),
                                   INLA = "https://inla.r-inla-download.org/R/stable"),
                 dependencies = TRUE)
```

# Acknowledgement

The `sfclust` package is based on the methodology and code from the paper *Bayesian
clustering of spatial functional data with application to a human mobility study during
COVID-19*. We acknowledge the authors for their contributions and inspiration.
