---
title: "sfclust - Normal Distributed Data Clustering"
author: "Ruiman Zhong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. sfclust - Simulated Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
  body {
    font-family: Arial, sans-serif;
    margin: 0;
  }

  /* Main content that spans the full width */
  .main-content {
    margin: 0; /* Remove the margin for full-width content */
    padding: 20px; /* Optional padding for better readability */
    width: 100%; /* Take up 100% of the page width */
    box-sizing: border-box; /* Ensure padding doesn’t cause overflow */
  }

  h1, h2, h3 {
    margin-left: 0; /* Ensure headers are aligned properly */
  }

  .page-break {
    page-break-after: always;
  }
</style>

<div class="main-content">

## Installation
To install the development version of the `sfclust` package from GitHub, you can use the following command:
```{r eval=FALSE, warning=FALSE, include=FALSE}
# Install development version from GitHub
devtools::install_github("RuimanZhong/sfclust")
```
Additionally, the package relies on INLA (Integrated Nested Laplace Approximations) for its computations. To install INLA, run the following:
```{r eval = FALSE}
# Install INLA
install.packages("INLA", repos = c(getOption("repos"),
                                   INLA = "https://inla.r-inla-download.org/R/stable"),
                 dependencies = TRUE)
```
[Next: Simulation](#simulation)

<div class="page-break"></div>

## Data Preparation
<a id="simulation"></a>
We begin by loading the required packages. These packages will assist with data manipulation, visualization, and running the `sfclust` algorithm. We load the a sftime dataset called `simu_data`. In the 3rd vignette, we introduce how to create a valid `stars` object as the input of the algorithm.

```{r warning=FALSE, include=FALSE}
library(igraph)
library(sf)
library(sfclust)
library(INLA)
library(ggplot2)
library(stars)
library(dplyr)
data("simu_data")
map <- st_as_sf(st_get_dimension_values(stdata, "geometry"))
```

[Next: Modeling](#modeling)

<div class="page-break"></div>

## Modeling
<a id="modeling"></a>

In this section, we show how to use `sfclust()` algorithm to find the spatial clusters. First we will create a stars object.
In this example, we assume the likelihood is normal. The latent effects
$f_{c}$ for $c = c,1, ..., c,n$ are represented with a random walk process, imposing the conditions $f_{c,t} - f_{c,t-1} \sim N(0, \nu^{-1})$ for t = 2,...,n. The model is specified as

$$
Y_{it} \mid \mu_{it}, \boldsymbol{\theta}_{T} \stackrel{ind}{\sim} \text{Normal}(\mu_{it},\epsilon^2_{it} ),
$$

$$
\mu_{it} = \alpha_{c_i} + f_{c_i,t} + \epsilon_{it},
$$
$$
\pi(\boldsymbol{f}_c | \nu_c, \boldsymbol{\theta}_{T}) \propto \nu_c^{(n-1)/2}exp(-\frac{\nu_c}{2}\boldsymbol{f}_c^{T}\boldsymbol{S}_{f}\boldsymbol{f}_c)
$$
$$
\epsilon_{it} \mid \boldsymbol{\theta}_{T} \sim N(0, 1/\tau_{c_i}^2),
$$
where $\boldsymbol{S}_f$ is the structure matrix obtained from the imposed conditions. The prior for the hyper-parameter $\nu_c$ is imposed as follows
$$
log(\nu_c) \sim LogGamma(1, 10^{-5})
$$

Next, we will show how to implement our clustering algorithm step by step. 

To facilitate the analysis of spatial data through clustering and network analysis, we employ the `genclust()` function, which is designed to initialize a graph from spatial data and compute its MST. The primary objective is to generate a graph that captures the spatial relationships between regions, which can then be used for clustering.

The function takes as input an sf object representing the spatial data (`map`), which contains connected polygons as areal data. In the function, we use adjacency matrix to construct the graph. The MST is computed from the constructed graph, which serves as the foundation for defining clusters within the spatial domain. The parameter `weights` is related to the edges weights to create MST T, by default it is `weights = NULL`, indicating that weights are generated from U[0,1].  

For instance, in our analysis, we utilize the following function call to construct the initial graph and compute the MST:


```{r echo=TRUE, warning=FALSE}
# Build MST 
set.seed(43)
# create graph
graphdata <- genclust(map, nclust = 15)
```

In this example, with the number of clusters (`nclust`) set to 15. The output, geodata, contains both the initial graph $G$ (graph0) and the MST $T$ ($mstgraph\_ini$), which are essential for subsequent clustering steps.

The following example demonstrates how to use the `sfclust()` function to perform spatial functional clustering with a normal likelihood model. The number of the clusters follows 
$$
\pi(C = n_{init})(1-c)^{n_{init}}
$$
In our algorithm, we will automatically create the three indexes: $idt$ represents the temporal index, $ids$ represents the spatial index, and $id$ represents the total data size. We could use those indexes without specifying them in the formula. The function `data_all()` shows the constructed data.
```{r}
head(data_all(stdata))
```

## Clustering Experiment

In this experiment, the temporal domain is modeled using a first-order random walk (rw1) to capture the smooth temporal trends. The formula for the model is defined as follows:
```{r}
formula <- Y ~ 1 + f(idt, model = "rw1")
```

We executed our clustering algorithm with 2000 iterations on the simulated data. The data is stored as a `stars` object, with `stnames = c("geometry", "time")` specifying the spatial (`geometry`) and temporal (`time`) dimensions of the data. The algorithm incorporates both spatial and temporal information to perform clustering.

We set the `move_prob` hyperparameter to specify probabilities for the four RJMCMC steps: birth, death, change, and hyperparameter updates. The values are defined as `move_prob = c(0.425, 0.425, 0.1, 0.05)`, representing:
- 42.5% probability for birth moves.
- 42.5% probability for death moves.
- 10% probability for change moves.
- 5% probability for hyperparameter updates.

Additionally, the penalty parameter \( q \) was set to 0.5, balancing model complexity and data fit.

The `sfclust()` function was used to fit the model to the data with the following settings:
- `correction = T`: Enables a correction mechanism for clustering for random walk model.
- `niter = 2000`: Specifies the total number of MCMC iterations.
- `burnin = 0`: No burn-in iterations are discarded.
- `thin = 1`: Retains every iteration without thinning.
- `formula = formula`: The defined random walk model `Y ~ 1 + f(idt, model = "rw1")` accounts for temporal dependencies.

The code used to run the experiment is shown below:

```{r, eval=FALSE}
set.seed(1234)
result <- sfclust(stdata, graphdata = graphdata,stnames = c("geometry", "time"),
                  move_prob = c(0.425, 0.425, 0.1, 0.05),q = 0.5, correction = T, 
                  niter = 2000, burnin = 0, thin = 1, formula = formula ) 

```{r include=FALSE}
result <- readRDS('~/Documents/Project4/normal_2001.rds')
```

## Results illustration

In this section, we illustrate the clustering results using the `summary()` and `plot()` functions. These tools provide comprehensive insights into the clustering process and outcomes. The `summary()` function summarizes key details at a specific iteration of the MCMC chain, such as cluster assignments, the size of each cluster, within-cluster models, and the total log marginal likelihood. The `plot()` function complements this by visually presenting the spatial cluster partition, temporal mean functions for selected clusters, and the convergence of the log marginal likelihood, offering both quantitative and qualitative perspectives on the clustering results. Together, these functions facilitate a thorough examination of the model's performance and clustering structure.


```{r}
summary(result)
```
The function shows the clusters, the number of elements per cluster, the within-cluster model, and total log marginal likelihood. This summary allows you to evaluate the clustering configuration at different stages of the MCMC process.

The plot() function offers a detailed visualization of clustering results, generating three plots: the spatial partition of clusters, the mean functions of clusters in the temporal domain, and the convergence of log marginal likelihood over iterations. To visualize the results at a specific MCMC iteration, the sample parameter can be set to the desired iteration index. For example, setting sample = 2000 will display the results from the 2000th iteration.

When the clustering process results in a large number of clusters, the clusters parameter allows you to select specific clusters for visualization by providing their labels. For instance, specifying `clusters = 1:5` focuses the plot on clusters 1 to 5. Additionally, the `sort = TRUE` parameter sorts cluster labels in descending order of the number of members, helping to prioritize the largest clusters in the visualization.

An example workflow would involve running the MCMC chain and plotting results at a specific iteration. For instance, to visualize iteration 2000 and plot only clusters 1 to 5 sorted by size, use the following code:
```{r fig.height=4, fig.width=8, warning=FALSE}
plot(result, sample = 2000, clusters = 1:5, sort = TRUE)
```


[Next: US COVID Data](US_Covid.html)

</div>
<div class="sidebar">
  <a href="#installation">Installation</a>
  <a href="#simulation">Simulation</a>
  <a href="#modeling">Modeling</a>

</div>
