---
title: "Advanced features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.height = 6.5, fig.width = 7.5, dpi = 90, out.width = '100%')
knitr::opts_chunk$set(comment = "#>")
```

```{r include = FALSE}
path_figures <- here::here(file.path("tools", "figures"))
save_figures <- dir.exists(path_figures)
```

In this vignette we present additional features of the `sfclust` package.

## Packages

We begin by loading the required packages.

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

## Data

The simulated dataset used in this vignette, `stgaus`, is included in our package. It is
a `stars` object with one variable, `y`, and two dimensions: `geometry` and `time`. The
dataset represents continuous Gaussian measurements from 100 regions, observed daily over
91 days, starting in January 2024.

```{r}
data("stgaus")
stgaus
```

To gain some initial insights, we can aggregate the data weekly:

```{r, fig.height = 5}
stweekly <- aggregate(stgaus, by = "week", FUN = mean)
ggplot() +
    geom_stars(aes(fill = y), stweekly) +
    facet_wrap(~ time, ncol = 5) +
    scale_fill_distiller(palette = "RdBu") +
    theme_bw()
```

We can also examine the trends for each region:

```{r, fig.height = 6}
stgaus |>
    st_set_dimensions("geometry", values = 1:nrow(stgaus)) |>
    as_tibble() |>
    ggplot() +
    geom_line(aes(time, y, group = geometry, color = factor(geometry)), linewidth = 0.3) +
    theme_bw() +
    theme(legend.position = "none")
```

Some regions exhibit similar trends over time, but the overall patterns are more complex
than polynomial functions. Our goal is to cluster these regions while accounting for
spatial contiguity.

## Clustering

### Model

We assume that the response variable \( Y_{it} \) for region \( i \) at time \( t \) follows a normal distribution given the partition \( M \):
$$
Y_{it} \mid \mu_{it}, M \stackrel{ind}{\sim} \text{Normal}(\mu_{it}, \sigma^2),
$$
where the mean \( \mu_{it} \) is modeled based on the cluster assignment \( c_i \):
$$
\mu_{it} = \alpha_{c_i} + f_{c_i,t},
$$
where \( \alpha_{c_i} \) is the intercept for cluster \( c_i \), and \( f_{c_i,t} \) is a latent random effect modeled as a random walk process. Specifically, we impose the condition:
$$
f_{c_i,t} - f_{c_i,t-1} \sim N(0, \nu^{-1}), \quad \text{for } t = 2, \dots, n.
$$

The prior for the hyperparameter \( \nu_c \) is defined as:
$$
\log(\nu_c) \sim \text{Normal}(-2, 1).
$$

### INLA integration

`sfclust` integrates seamlessly with `inla()`, letting us specify the within-cluster model
above exactly as we would in a standard `INLA` workflow. The model formula (including
random effects and priors) and any additional arguments are passed directly to `inla()`.
The following formula defines a linear predictor with a first-order random walk and the
prior `Normal(-2, 1)` on the log-precision described above; we reuse this `formula` object
in the examples below.

```{r}
formula <- y ~ f(id_time, model = "rw1",
  hyper = list(prec = list(prior = "normal", param = c(-2, 1))))
```

You can define other latent models and priors following `INLA` conventions. Further
documentation is available at [https://www.r-inla.org/documentation](https://www.r-inla.org/documentation).

### Penalizing the number of clusters

The `logpen` argument sets the log-penalty on the number of clusters, specifically
`log(1 - q)`, where `q` is the prior probability of keeping a cluster: there is no penalty
when `logpen = 0`, and more negative values increasingly favor fewer clusters. A large
negative value such as `-50` acts as a strong parsimony prior, meaning that any proposal
to increase the number of clusters must improve the log marginal likelihood by around 50
units to have a high acceptance probability.

### Save options

Besides `niter`, `burnin` discards the first iterations before saving samples, and `thin`
retains only every `thin`-th iteration, reducing autocorrelation in the chain. Note that
`niter` counts iterations *after* burn-in, so the algorithm runs `burnin + niter`
iterations in total. To reduce the risk of losing intermediate results, especially given
the computational cost of the algorithm, you can use the `path_save` argument to specify a
file for saving the output, and `nsave` to define the frequency at which samples are
saved.

### Sampling with `sfclust`

We now put all of the above together. We start with 20 initial clusters (`nclust = 20`)
and run the algorithm for only 50 iterations, using the strong penalty and formula defined
above, and saving intermediate results to `"stgaus-mcmc-initial.rds"`.

```{r, eval = FALSE}
set.seed(123)
result0 <- sfclust(stgaus, nclust = 20, formula = formula, logpen = -50,
  niter = 50, burnin = 10, thin = 2, nmessage = 10,
  path_save = "stgaus-mcmc-initial.rds")
result0
```

```{r, echo = FALSE}
result0 <- readRDS(system.file("vigdata", "gaussian-mcmc1.rds", package = "sfclust"))
result0
```

The output indicates that after starting with 20 clusters, the algorithm created 11 new
clusters (births), removed 8 clusters (deaths), changed the membership of 1 cluster, and
modified the minimum spanning tree (MST) 2 times.

The `plot` method allows us to select which graph to produce. For example, we can
visualize only the log marginal likelihood to diagnose convergence. With just 50
iterations, we can see that the log marginal likelihood has not yet achieved convergence.

```{r, fig.dpi = 72, fig.height = 6}
plot(result0, which = 3)
```

### Continue sampling

To achieve convergence, we can continue the sampling using the `update` function and
specify the number of new iterations with the `niter` argument:

```{r, eval = FALSE}
result <- update(result0, niter = 1000, nsave = 500, path_save = "stgaus-mcmc.rds")
result
```

```{r, echo = FALSE}
result <- readRDS(system.file("vigdata", "gaussian-mcmc2.rds", package = "sfclust"))
result
```

With 1000 additional iterations, there have been many clustering movements. Furthermore,
when visualizing the results, we can observe that the log marginal likelihood has achieved
convergence.

```{r, fig.dpi = 72, fig.height = 6}
plot(result, which = 3)
```

The side-by-side comparison below highlights the contrast: panel (A) shows the unstable
log marginal likelihood from the short initial run, while panel (B) shows the stabilized
trajectory after the full run, indicating convergence.

```{r, fig.height = 3.5}
gg1 <- plot(result0, which = 3) + labs(subtitle = "(A)")
gg2 <- plot(result, which = 3) + labs(subtitle = "(B)")
gg1 + gg2
if (save_figures) {
  ggsave(file.path(path_figures, "stgaus-marginal-likelihood.pdf"), width = 10, height = 4.5,
    device = cairo_pdf)
}
```

The `update` function can also be used to select a different past iteration as the active
clustering, without running additional MCMC. This is useful when you want to inspect or
use a specific sample rather than the last one:

```{r}
result_other <- update(result, sample = 750)
result_other
```

## Results

The final iteration indicates that the algorithm identified 12 clusters, with a log
marginal likelihood of 12,602.37. The largest cluster consists of 27 regions, while the
smallest contains only one region.

```{r}
summary(result, sort = TRUE)
```

Let's visualize the regions grouped by cluster.

```{r, fig.height = 4}
plot(result, which = 1:2, sort = TRUE, legend = TRUE)
```

Let's visualize the original data grouped by cluster.

```{r, fig.height = 5}
plot_clusters_series(result, y, sort = TRUE) +
  facet_wrap(~ cluster, ncol = 5) +
  labs(title = "Risk per cluster", y = "Response")
```

