---
title: "Weekly relative risk of COVID-19 in US states"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weekly relative risk of COVID-19 in US states}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", eval = TRUE)
knitr::opts_chunk$set(fig.height = 6.5, fig.width = 7.5, dpi = 100, out.width = '100%')
knitr::opts_chunk$set(comment = "#>")
```

```{r include = FALSE}
path_figures <- here::here(file.path("tools", "figures"))
save_figures <- dir.exists(path_figures)
```

In this vignette, we use `sfclust` to identify US states with similar weekly relative risk
of COVID-19 during 2020.

## Load packages and data

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

The data for this application is available on our GitHub repository:
[https://github.com/ErickChacon/sfclust/blob/main/tools/data/usacovid.rds](https://github.com/ErickChacon/sfclust/blob/main/tools/data/usacovid.rds).
It contains weekly data for 49 states, including the number of cases, the population, and
the number of expected cases.

```{r}
link <- "https://github.com/ErickChacon/sfclust/blob/main/tools/data/"
usacovid <- readRDS(gzcon(url(paste0(link, "usacovid.rds?raw=true"))))
usacovid
```

Note that the dimension names are `space` and `time`. Since the spatial dimension is not
the default `"geometry"`, you must specify `spnames = "space"` when using `sfclust()`.

## Exploratory analysis

We begin by visualizing the weekly relative risk of COVID-19 across the 49 states. Higher
risk levels are observed toward the end of the year.

```{r, fig.height = 6.5}
ggplot() +
    geom_stars(aes(fill = cases/expected), data = usacovid) +
    facet_wrap(~ time, ncol = 7) +
    scale_fill_distiller(palette = "RdBu") +
    labs(fill = "Relative risk") +
    theme_bw(base_size = 7) +
    theme(legend.position = "bottom")
```

We can also examine the time series of relative risk for each state. Peaks are evident
around April, July, and especially December.

```{r, fig.height = 5}
usacovid |>
    st_set_dimensions("space", values = 1:ncol(usacovid)) |>
    as_tibble() |>
    ggplot() +
    geom_line(aes(time, cases/expected, group = space, color = factor(space)), linewidth = 0.3) +
    geom_point(aes(time, cases/expected, group = space, color = factor(space))) +
    labs(y = "Relative risk", x = NULL) +
    theme_bw() +
    theme(legend.position = "none")
```

## Spatial clustering

### Model fitting

We assume that the logarithm of the relative risk can be explained by:

1. A polynomial trend over `time`
2. An autoregressive effect over `id_time`
3. An unstructured random effect across states and times (`id`)

We begin with one cluster per state (49 clusters total). Be sure to set the spatial
dimension name: `spnames = "space"`. Two additional arguments deserve mention: `nsave`
periodically saves intermediate results to disk every 1000 iterations (useful for long
runs); `control.inla` disables the variational Bayes approximation (`control.vb`) and
uses empirical Bayes integration (`int.strategy = "eb"`) for numerical stability with
this model.

```{r, eval = FALSE}
set.seed(123)
formula <- cases ~ 1 + poly(time, 3) + f(id_time, model = "ar1") + f(id)
result <- sfclust(usacovid, nclust = 49, spnames = "space",
    formula = formula, family = "poisson", E = expected,
    niter = 4000, burnin = 0, thin = 10, nmessage = 10, nsave = 1000,
    control.inla = list(control.vb = list(enable = FALSE), int.strategy = "eb"))
result
```

```{r, echo = FALSE}
formula <- cases ~ 1 + poly(time, 3) + f(id_time, model = "ar1") + f(id)
result <- readRDS(here::here(file.path("tools", "data", "usacovid-mcmc.rds")))
result
```

Summary of clustering steps:

* 20 cluster splits
* 57 cluster merges
* 13 cluster composition changes
* 223 updates to the minimum spanning tree

400 samples were retained (after thinning) from 4000 iterations. The final marginal likelihood was -19199.61.

### Results

```{r}
summary(result, sort = TRUE)
```

The `summary()` output shows that 7 out of 12 clusters contain more than one state. The
two largest clusters each include 9 states, while the remaining clusters are smaller. In
order to verify the adequacy of this clustering, we check the convergence using the
`plot()` function with option `which = 3`.

```{r, fig.height = 4.5}
plot(result, which = 3)
```

The figure indicates that the log marginal likelihood improves significantly within the
first 100 samples (after thinning) and stabilizes afterward from sample 200. Now, we can
visualize the spatial cluster assignment and the predicted mean for each cluster:

```{r, fig.height = 3}
plot(result, which = 1:2, legend = TRUE, sort = TRUE)
```

```{r, echo = FALSE}
if (save_figures) {
  ggsave(file.path(path_figures, "usacovid-plot.pdf"), width = 10, height = 3,
    device = cairo_pdf)
}
```

The `plot()` output indicates that:

* The largest cluster is in the southwestern US, followed by clusters in central and northwestern regions.
* Mean relative risk increases gradually across the year, with different peaks per cluster.

The `fitted` function returns a `stars` object containing prediction summaries: the
`cluster` assignment, the `mean_cluster` linear predictor, and its inverse
(`mean_cluster_inv`).

```{r}
us_fit <- fitted(result, sort = TRUE)
us_fit
```

### Empirical risk per cluster

We use the cluster assignments to analyze and visualize the empirical relative risk per
cluster. This can easily be done with the `plot_clusters_series()` function:

```{r, fig.height = 3.5}
plot_clusters_series(result, cases/expected, sort = TRUE, clusters = 1:12) +
  facet_wrap(~ cluster, ncol = 5) +
  scale_x_date(date_breaks = "2 months", date_labels =  "%b") +
  labs(y = "Relative risk")
```

```{r, echo = FALSE}
if (save_figures) {
  ggsave(file.path(path_figures, "usacovid-risk-per-cluster.pdf"),
    width = 10, height = 4, device = cairo_pdf)
}
```

This figure shows distinct epidemic dynamics by cluster:

* Cluster 1 shows an increasing risk with a peak in July–August.
* Cluster 2 peaks around November–December, with a decline by year-end.
* Cluster 3 shows early activity around March and high risk near January.
* Clusters 4–7 follow various intermediate trends.
* Clusters 8–12 exhibit less typical behavior.

To complement the empirical view, we can also visualize the cluster-level mean log
relative risk over time using the `fitted` function with `aggregate = TRUE`. This returns
a `stars` object with one geometry per cluster, making it straightforward to produce a
faceted map of the smoothed cluster predictions:

```{r, fig.height = 6.5}
pred <- fitted(result, sort = TRUE, aggregate = TRUE)
ggplot() +
    geom_stars(aes(fill = mean_cluster), data = pred) +
    facet_wrap(~ time, ncol = 7) +
    scale_fill_distiller(palette = "RdBu") +
    labs(title = "Weekly risk", fill = "Risk") +
    theme_bw(base_size = 7) +
    theme(legend.position = "bottom")
```

<!-- vim: set softtabstop=2 sw=2 tabstop=2 expandtab: -->
