---
title: "Covid-19 in US states during 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid-19 in US states during 2020}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", eval = TRUE)
knitr::opts_chunk$set(fig.height = 6.5, fig.width = 7.5, dpi = 90, out.width = '100%')
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
```

In this vignette, we use `sfclust` to identify US states with similar relative risk of
Covid-19 during 2020.

## Load packages and data

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

The data for this application is available on our GitHub repository:
[https://github.com/ErickChacon/sfclust/blob/main/tools/data/usacovid.rds](https://github.com/ErickChacon/sfclust/blob/main/tools/data/usacovid.rds).
It contains weekly data for 49 states, including the number of cases, the population, and
the number of expected cases.

```{r}
usacovid <- readRDS(here::here(file.path("tools", "data", "usacovid.rds")))
usacovid
```

Note that the dimension names are `space` and `time`. Since `sfclust` expects them to be
`c("geometry", "time")`, you must specify the `stnames` argument when using the function
`sfclust()` of the package.

## Exploratory analysis

We begin by visualizing the weekly relative risk of Covid-19 across the 49 states. Higher
risk levels are observed toward the end of the year.

```{r, fig.height = 6}
ggplot() +
    geom_stars(aes(fill = cases/expected), data = usacovid) +
    facet_wrap(~ time) +
    scale_fill_distiller(palette = "RdBu") +
    labs(fill = "Relative risk") +
    theme_bw() +
    theme(legend.position = "bottom")
```

We can also examine the time series of relative risk for each state. Peaks are evident
around April, July, and especially December.

```{r, fig.height = 5}
usacovid |>
    st_set_dimensions("space", values = 1:ncol(usacovid)) |>
    as_tibble() |>
    ggplot() +
    geom_line(aes(time, cases/expected, group = space, color = factor(space)), linewidth = 0.3) +
    geom_point(aes(time, cases/expected, group = space, color = factor(space))) +
    labs(y = "Relative risk", x = NULL) +
    theme_bw() +
    theme(legend.position = "none")
```

## Spatial clustering

### Model fitting

We assume that the logarithm of the relative risk can be explained by:

1. A polynomial trend over `time`
2. An autoregressive effect over `idt`
3. An unstructured random effect across states and times (`id`)

We begin with one cluster per state (49 clusters total). Be sure to set the dimension
names: `stnames = c("space", "time")`.

```{r, eval = FALSE, echo = FALSE}
# formula <- cases ~ 1 + poly(time, 3) + f(idt, model = "ar1") + f(id)
# geodata <- genclust(st_geometry(usacovid), nclust = 49)
#
# set.seed(123)
# result <- sfclust(usacovid, graphdata = geodata, stnames = c("space", "time"),
#     formula = formula, family = "poisson", E = expected,
#     niter = 4000, burnin = 0, thin = 10, nmessage = 10, nsave = 100,
#     path_save = file.path("tools", "data", "usacovid-mcmc.rds"))
```

```{r, eval = FALSE}
formula <- cases ~ 1 + poly(time, 3) + f(idt, model = "ar1") + f(id)
geodata <- genclust(st_geometry(usacovid), nclust = 49)

set.seed(123)
result <- sfclust(usacovid, graphdata = geodata, stnames = c("space", "time"),
    formula = formula, family = "poisson", E = expected,
    niter = 4000, burnin = 0, thin = 10, nmessage = 10, nsave = 100,
    path_save = "usacovid-mcmc.rds")
result
```

```{r, echo = FALSE}
formula <- cases ~ 1 + poly(time, 3) + f(idt, model = "ar1") + f(id)
geodata <- genclust(st_geometry(usacovid), nclust = 49)
result <- readRDS(here::here(file.path("tools", "data", "usacovid-mcmc.rds")))
result
```

Summary of clustering steps:

* 36 cluster splits
* 74 cluster merges
* 12 cluster composition changes
* 187 updates to the minimum spanning tree

400 samples were retained (after thinning) from 4000 iterations. The final marginal likelihood was -19196.77.

### Results

```{r}
summary(result, sort = TRUE)
```

The `summary()` output shows that 7 out of 11 clusters contain more than one state. The
largest cluster includes 10 states, while the second includes 9, and so on. In order to
verify the adequacy of this clustering, we check the convergence using the `plot()`
function with option `which = 3`.

```{r}
plot(result, which = 3)
```

The figure indicates that the marginal likelihood improves significantly within the first
100 iterations (after thinning) and stabilizes afterward from sample 200. Now, we can
visualize the spatial cluster assignment and the predicted mean for each cluster:

```{r, fig.height = 3}
plot(result, which = 1:2, legend = TRUE, sort = TRUE)
```

```{r, echo = FALSE}
path_figures <- here::here(file.path("tools", "figures"))
ggsave(file.path(path_figures, "usacovid-plot.pdf"), width = 10, height = 3, device = cairo_pdf)
```

The `plot()` output indicates that:

* The largest cluster is in the southwestern US, followed by clusters in central and northwestern regions.
* Mean relative risk increases gradually across the year, with different peaks per cluster.

The `fitted` function return an `stars` object with prediction summaries after fitting,
the `cluster` assignment, the `mean_cluster` linear predictor, the inverse of the linear
predictor (`mean_cluster_inv`).

```{r}
us_fit <- fitted(result, sort = TRUE)
us_fit
```

### Empirical risk per cluster

We use the cluster assignments to analyze and visualize the empirical relative risk per cluster. This can easily be done with the `plot_cluster_series()` function:

```{r, fig.height = 5}
plot_clusters_series(result, cases/expected) + labs(y = "Relative risk")
```

```{r, echo = FALSE}
ggsave(file.path(path_figures, "usacovid-risk-per-cluster.pdf"), width = 10,
  height = 7, device = cairo_pdf)
```

This figure shows distinct epidemic dynamics by cluster:

* Cluster 1 shows an increasing risk with a peak in July–August.
* Cluster 2 peaks around November–December, with a decline by year-end.
* Cluster 3 shows early activity around March and high risk near January.
* Clusters 4–7 follow various intermediate trends.
* Clusters 8–11 exhibit less typical behavior.

```{r, echo = FALSE}
# pred <- fitted(result, sort = TRUE, aggregate = TRUE)
# ggplot() +
#     geom_stars(aes(fill = mean_cluster), data = pred) +
#     facet_wrap(~ time) +
#     scale_fill_distiller(palette = "RdBu") +
#     labs(title = "Daily risk", fill = "Risk") +
#     theme_bw(base_size = 7) +
#     theme(legend.position = "bottom")
```

```{r, echo = FALSE}
# # saveRDS(list(x = 1, b = 2), "find-problem.rds")
#
#  # -19550.807576937
# plot(result)
#  # -19449.8533593845
#  # -19306.77
#  # -19284.15
#
# # result2 =  update(result, niter = 2000, path_save = file.path("inst", "vigdata", "usacovid2.rds"))
#
# result = readRDS("clust_usacovid1.rds")
# # result2 =  readRDS(file.path("inst", "vigdata", "usacovid2.rds"))
# summary(result)
# # summary(result2)
# # library(sfclust)
# plot(result, which = 1)
```

<!-- vim: set softtabstop=2 sw=2 tabstop=2 expandtab: -->
