---
title: "Clustering raster spatio-temporal data (NDWI2)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering raster spatio-temporal data (NDWI2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", eval = TRUE)
knitr::opts_chunk$set(fig.height = 5, fig.width = 7.5, dpi = 120, out.width = '100%')
knitr::opts_chunk$set(comment = "#>")
```

```{r include = FALSE}
path_figures <- here::here(file.path("tools", "figures"))
save_figures <- dir.exists(path_figures)
```

In this vignette, we use the `sfclust` package with **raster** spatio-temporal
data. We cluster the functional trend of the NDWI2 vegetation water index (stored as
the `ndwi2` column) for each cell in the agricultural experiment in "El Chaparrillo"
over the period May–December 2025. The `ndwi2` values are derived from Sentinel-2
imagery and are bundled with the package as the `chapa` dataset.

## Load packages and data

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

```{r}
chapa
```

## Exploratory analysis

We start by visualizing the spatio-temporal behaviour of NDWI2.

```{r, fig.height = 8}
ggplot() +
  geom_stars(aes(fill = ndwi2), data = chapa) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "grey50") +
  labs(fill = "NDWI2") +
  theme_void(base_size = 7) +
  theme(legend.position = "bottom")
```

Notice that these data contain cells with NA values. `sfclust()` automatically builds a
graph that respects the connectivity of the cells and excludes the NA cells from the
clustering.

We can also plot the pixel-level time series to see how NDWI2 evolves over the period.

```{r, fig.height = 4}
chapa_aux <- as_tibble(chapa) |>
  filter(!is.na(ndwi2)) |>
  mutate(pixel = paste(x, y, sep = "-")) |>
  group_by(pixel) |>
  mutate(pixel_mean = mean(ndwi2)) |>
  ungroup()

ggplot(chapa_aux, aes(time, ndwi2, group = pixel)) +
  geom_line(aes(color = pixel_mean), alpha = 0.5, linewidth = 0.5) +
  scale_color_gradient2(low = 2, mid = "gray30", high = 4, midpoint = 0.15) +
  labs(title = "NDWI2 time series per pixel", x = NULL, y = "NDWI2", color = "Mean") +
  theme_bw()
```

## Model fitting

We model NDWI2 within each cluster as a quadratic polynomial in time (cell $i$ and time
$t$):

$$\text{NDWI2}_{it} = \beta_0 + \beta_1 t^{(1)} + \beta_2 t^{(2)} + \varepsilon_{it}$$

where $t^{(1)}$ and $t^{(2)}$ are polynomial basis functions with respect to `time` and error term
$\varepsilon_{it}$  normally distributed with zero-mean and variance $\sigma^2$.

```{r, eval = FALSE}
set.seed(7)
result <- sfclust(
    chapa, nclust = 335, formula = ndwi2 ~ poly(time, 2),
    niter = 4000, thin = 10, nmessage = 100, nsave = 1000,
    path_save = "chapa-mcmc.rds"
)
result
```

```{r, echo = FALSE}
result <- readRDS(here::here(file.path("tools", "data", "chapa-mcmc.rds")))
result
result <- update(result)
```

Over the course of sampling, 40 cluster additions, 329 merges, 43 reassignments, and 212
hyperparameter updates occurred. The marginal likelihood reached a value of 46,972.3.
After thinning, 400 samples were retained.

## Results

According to the `summary()` output, 13 out of 46 clusters have more than 8 members. The
largest clusters contain 32, 28, and 23 pixels, respectively.

```{r}
summary(result, sort = TRUE)
```

The figure below shows the spatial distribution of the inferred clusters over the raster
grid, the fitted functional shapes, and the traceplot of the log marginal likelihood
across MCMC samples. Convergence appears to have been practically achieved.

```{r, fig.height = 3.5}
plot(result)
```

Let's visualize only the 8 biggest clusters. Cluster 1 is located in the southeast and has
lower ndwi2 levels, while cluster 2 is located north of center and usually has higher
ndwi2 values.

```{r, fig.height = 3}
plot(result, which = 1:2, sort = TRUE, clusters = 1:8, legend = TRUE)
```

## Fitted values

We can visualize the cluster mean values over space and time. It shows that ndwi2 values
are lower around July, with higher values in November–December.

```{r, fig.height = 8}
pred <- fitted(result, sort = TRUE)
ggplot() +
  geom_stars(aes(fill = mean_cluster), data = pred) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "grey50") +
  labs(fill = "NDWI2") +
  theme_void(base_size = 7) +
  theme(legend.position = "bottom")
```

Finally, let's visualize the functional trends of the 8 biggest clusters. We can see that
the pixels belonging to each cluster actually have a very similar evolution over time.

```{r, fig.height = 5}
plot_clusters_series(result, ndwi2, sort = TRUE, clusters = 1:8)
```
