---
title: "Clustering raster spatio-temporal data (NDWI2)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering raster spatio-temporal data (NDWI2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", eval = TRUE)
knitr::opts_chunk$set(fig.height = 5, fig.width = 7.5, dpi = 90, out.width = '100%')
knitr::opts_chunk$set(comment = "#>")
```

In this vignette, we use the `sfclust` package with **raster** (gridded) spatio-temporal
data. Unlike the polygon- or point-based examples in other vignettes, here the spatial
units are regular grid cells identified by their `x` and `y` coordinates. This exercises
the raster support introduced via `sp_dims = c("x", "y")` and `raster_adjacency()`.

The dataset used is `stndwi`, a `stars` raster object containing the Normalized
Difference Water Index 2 (NDWI22) derived from Sentinel-2 satellite imagery over
El Chaparrillo, an agricultural research center in Spain, with 86 acquisition dates
spanning May–December 2025.

## Load packages and data

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

```{r}
data(stndwi)
# stndwi <- aggregate(stndwi, by = "week", FUN = function(x) mean(x, na.rm = TRUE))
stndwi
```

The raster has `r dim(stndwi)["x"]` columns and `r dim(stndwi)["y"]` rows, giving a
grid of `r prod(dim(stndwi)[c("x","y")])` cells in total, of which
`r sum(!is.na(stndwi[["ndwi2"]][,,1]))` are non-NA (the study area mask).

## Exploratory analysis

We start by visualizing NDWI2 at a few time points to get a sense of the spatial patterns.

```{r, fig.height = 6}
ggplot() +
  geom_stars(aes(fill = ndwi2), data = stndwi) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey90") +
  labs(title = "NDWI2 raster — selected time points", fill = "NDWI2") +
  theme_bw() +
  theme(legend.position = "bottom")
```

We can also plot the pixel-level time series to see how NDWI2 evolves over the season.

```{r, fig.height = 4}
as_tibble(stndwi) |>
  filter(!is.na(ndwi2)) |>
  mutate(pixel = paste(x, y, sep = "-")) |>
  ggplot(aes(time, ndwi2, group = pixel)) +
  geom_line(alpha = 0.15, linewidth = 0.3, color = "steelblue") +
  labs(title = "NDWI2 time series per pixel", x = "Date", y = "NDWI2") +
  theme_bw()
```

## Building the spatial graph

For raster data, `genclust()` automatically detects the two spatial dimensions, builds
the 4-connected (rook) adjacency via `raster_adjacency()`, and restricts the graph to
non-NA pixels before constructing the spanning tree.

```{r}
set.seed(111)
geodata <- genclust(stndwi, nclust = 20, spnames = c("x","y"), response = "ndwi2")
```

## Model fitting

We model NDWI2 within each cluster as a quadratic polynomial in time plus an
individual-level noise term per observation:

$$\text{NDWI2}_{it} = \beta_0 + \beta_1 t + \beta_2 t^2 + \varepsilon_{it}$$

where `time` refers to the actual acquisition dates (coerced to numeric by `poly()`)
and `id` is the flat space-time index used to account for individual observation noise.

```{r}
formula <- ndwi2 ~ poly(time, 2) + f(id, model = "iid")

result <- sfclust(stndwi, graphdata = geodata, formula = formula,
  spnames = c("x", "y"), logpen = -50,
  niter = 200, burnin = 0, thin = 1, nmessage = 5,
  control.inla = list(control.vb = list(enable = FALSE)))
```

## Results

```{r}
result
summary(result, sort = TRUE)
```

The traceplot below shows the log marginal likelihood across MCMC samples.

```{r, fig.height = 4}
plot(result)
plot(result, which = 1)
```

The map below shows the spatial distribution of the inferred clusters over the raster
grid.

```{r, fig.height = 4}
plot_clusters_map(result, sort = TRUE, legend = TRUE)
```

The fitted cluster functions show the estimated mean NDWI2 trajectory for each cluster.

```{r, fig.height = 4}
plot_clusters_fitted(result, sort = TRUE, legend = TRUE)
```

We can also inspect the fitted values on the raster grid at selected time points.

```{r, fig.height = 6}
pred <- fitted(result, sort = TRUE)
pred$time <- NULL
# idx  <- round(seq(1, dim(pred)["time"], length.out = 20))
ggplot() +
  geom_stars(aes(fill = mean_cluster), data = pred) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey90") +
  labs(title = "Fitted NDWI2 — selected time points", fill = "NDWI2") +
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom")
```

Finally, we overlay the observed pixel time series per cluster with the estimated cluster
mean.

```{r, fig.height = 4}
plot_clusters_series(result, ndwi2, sort = TRUE)
```
