---
title: "Clustering raster spatio-temporal data (NDWI)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering raster spatio-temporal data (NDWI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", eval = TRUE)
knitr::opts_chunk$set(fig.height = 5, fig.width = 7.5, dpi = 90, out.width = '100%')
knitr::opts_chunk$set(comment = "#>")
```

In this vignette, we use the `sfclust` package with **raster** (gridded) spatio-temporal
data. Unlike the polygon- or point-based examples in other vignettes, here the spatial
units are regular grid cells identified by their `x` and `y` coordinates. This exercises
the raster support introduced via `sp_dims = c("x", "y")` and `raster_adjacency()`.

The dataset used is `testdata`, a `stars` raster object containing the Normalized
Difference Water Index (NDWI) over a study area, with 102 irregular acquisition dates
spanning May–December 2025.

## Load packages and data

```{r, warning = FALSE, message = FALSE}
library(sfclust)
library(stars)
library(ggplot2)
library(dplyr)
```

```{r}
data(testdata)
# testdata <- aggregate(testdata, by = "week", FUN = function(x) mean(x, na.rm = TRUE))
testdata
```

The raster has `r dim(testdata)["x"]` columns and `r dim(testdata)["y"]` rows, giving a
grid of `r prod(dim(testdata)[c("x","y")])` cells in total, of which
`r sum(!is.na(testdata[["ndwi"]][,,1]))` are non-NA (the study area mask).

## Exploratory analysis

We start by visualizing NDWI at a few time points to get a sense of the spatial patterns.

```{r, fig.height = 6}
ggplot() +
  geom_stars(aes(fill = ndwi), data = testdata) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey90") +
  labs(title = "NDWI raster — selected time points", fill = "NDWI") +
  theme_bw() +
  theme(legend.position = "bottom")
```

We can also plot the pixel-level time series to see how NDWI evolves over the season.

```{r, fig.height = 4}
as_tibble(testdata) |>
  filter(!is.na(ndwi)) |>
  mutate(pixel = paste(x, y, sep = "-")) |>
  ggplot(aes(time, ndwi, group = pixel)) +
  geom_line(alpha = 0.15, linewidth = 0.3, color = "steelblue") +
  labs(title = "NDWI time series per pixel", x = "Date", y = "NDWI") +
  theme_bw()
```

## Building the spatial graph

For raster data, `genclust()` automatically detects the two spatial dimensions, builds
the 4-connected (rook) adjacency via `raster_adjacency()`, and restricts the graph to
non-NA pixels before constructing the spanning tree.

```{r}
set.seed(111)
geodata <- genclust(testdata, nclust = 10, sp_dims = c("x","y"))
```

## Model fitting

We model NDWI within each cluster as a quadratic polynomial in time plus an
individual-level noise term per observation:

$$\text{NDWI}_{it} = \beta_0 + \beta_1 t + \beta_2 t^2 + \varepsilon_{it}$$

where `time` refers to the actual acquisition dates (coerced to numeric by `poly()`)
and `id` is the flat space-time index used to account for individual observation noise.

```{r}
formula <- ndwi ~ poly(time, 2) + f(id, model = "iid")

result <- sfclust(testdata, graphdata = geodata, formula = formula,
  sp_dims = c("x", "y"), logpen = -50,
  niter = 5, burnin = 0, thin = 1, nmessage = 5,
  control.inla = list(control.vb = list(enable = FALSE)))
```

## Results

```{r}
result
summary(result, sort = TRUE)
```

The traceplot below shows the log marginal likelihood across MCMC samples.

```{r, fig.height = 4}
plot(result)
plot(result, which = 3)
```

The map below shows the spatial distribution of the inferred clusters over the raster
grid.

```{r, fig.height = 4}
plot_clusters_map(result, sort = TRUE, legend = TRUE)
```

The fitted cluster functions show the estimated mean NDWI trajectory for each cluster.

```{r, fig.height = 4}
plot_clusters_fitted(result, sort = TRUE, legend = TRUE)
```

We can also inspect the fitted values on the raster grid at selected time points.

```{r, fig.height = 6}
pred <- fitted(result, sort = TRUE)
pred$time <- NULL
# idx  <- round(seq(1, dim(pred)["time"], length.out = 20))
ggplot() +
  geom_stars(aes(fill = mean_cluster), data = pred) +
  facet_wrap(~ time) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "grey90") +
  labs(title = "Fitted NDWI — selected time points", fill = "NDWI") +
  theme_bw(base_size = 7) +
  theme(legend.position = "bottom")
```

Finally, we overlay the observed pixel time series per cluster with the estimated cluster
mean.

```{r, fig.height = 4}
plot_clusters_series(result, ndwi, sort = TRUE)
```
